---
title: "SQL Fundamentals Workshop"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: false
runtime: shiny_prerendered
description: >
  Learn how to write tutorials with interactive SQL exercises.
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(WDI)
library(countrycode)
library(RSQLite)
library(DBI)
if (!rlang::is_installed("gradethis", version = "0.2.6.9000")) {
  stop("Please install `gradethis` from GitHub (rstudio/gradethis) for this example.")
}
library(gradethis)

wdi <- WDI::WDI(country = "all", start=2023, end=2023, extra=TRUE, indicator="SP.POP.TOTL") %>%
  filter(region != "Aggregates") %>%
  rename(population = `SP.POP.TOTL`) %>%
  select(country:latitude & !status)

db_con <- DBI::dbConnect(RSQLite::SQLite(), dbname = ":memory:")
DBI::dbWriteTable(db_con, "tblCountries", wdi)

tutorial_options(exercise.eval = FALSE)
new_recorder <- function(tutorial_id, tutorial_version, user_id, event, data) {
  cat(tutorial_id, user_id, event, data$sectionId, data$label, data$answers, data$correct, "\n", sep = "|")
}

options(tutorial.event_recorder = new_recorder)
```

## Basic SELECT syntax

The most common type of SQL operation you will perform in this class is
to retrieve data from a table or set of tables. This kind of operation,
known as a **query**, involves writing a structured code statement that
is broken down into elements known as **clauses**, and each clause is
initiated with a keyword.

For retrieving data, there are two clauses that are required: `SELECT`
and `FROM`, and they are written in that order:

``` markdown
SELECT
  column1,
  column2
FROM 
  someTable;
```

For this exercise, we're going to use a sample database with a single
table named `tblCountries` with the following columns:

| index | columnName  | columnDescription                                |
|:------|:------------|:-------------------------------------------------|
| 0     | country     | The name of the country                          |
| 1     | iso2c       | The country's 2-digit ISO code                   |
| 2     | iso3c       | The country's 3-digit ISO code                   |
| 3     | year        | The year for which the data was gathered         |
| 4     | population  | Total country population                         |
| 6     | lastupdated | The date on which the data was last updated      |
| 7     | region      | The world region in which the country is located |
| 8     | capital     | The country's capital city                       |
| 9     | longitude   | Longitude of the country's geographic center     |
| 10    | latitude    | Latitude of the country's geographic center      |

## SELECT

### SELECT exercise 1

Your goal is to query the `tblCountries` table and return only the
columns named `country`, `region`, `capital`, and `population`.

```{sql db-exSelect1, exercise = TRUE, connection = "db_con", output.var="my_result", exercise.reveal_solution = FALSE}

```

```{sql db-exSelect1-solution}
SELECT
  country,
  region,
  capital,
  population
FROM 
  tblCountries;
```

```{r db-exSelect1-check}
grade_this({
  .solution <- DBI::dbGetQuery(db_con, .solution_code)
  
  cols_expected <- c("country", "region", "capital", "population")

  for (column in cols_expected) {
    if (!column %in% names(.result)) {
      fail("Did you forget to include `{column}`?")
    }
  }

  extra_cols <- setdiff(names(.result), cols_expected)
  if (length(extra_cols)) {
    extra_cols <- knitr::combine_words(extra_cols, and = " or ", before = "`")
    fail("You don't need the columns {extra_cols}.")
  }

  pass_if_equal()
  fail()
})
  
```

## ORDER BY

It's often desirable to arrange (or sort) the data in a specific order.
This can help us with analyzing the data, finding anomalous rows we
might want to filter out, or prepping it for other downstream activities
like visualizing it or loading it into another system.

The `ORDER BY` clause is what we can use to sort our results by a
column, or a list of columns.

``` markdown
SELECT
  column1,
  column2
FROM 
  someTable
ORDER BY
  column1;
```

### ORDER BY exercise 1

Using the same query as before, arrange the results by `population`.

```{sql db-exOrderBy1, exercise = TRUE, connection = "db_con", output.var="my_result", exercise.reveal_solution = FALSE}
SELECT
  country,
  region,
  capital,
  population
FROM 
  tblCountries;
  
```

```{sql db-exOrderBy1-solution}
SELECT
  country,
  region,
  capital,
  population
FROM 
  tblCountries
ORDER BY
  population;
```

```{r db-exOrderBy1-check}
grade_this({
  .solution <- DBI::dbGetQuery(db_con, .solution_code)
  # # if (unique(.solution$year) == "2023") {
  # #   fail("Did you remember to select only data from the year 2023?")
  # # }

  pass_if_equal()
  fail()
})

  
```

### ORDER BY exercise 2

In the last example, you noticed that using `ORDER BY population` caused
the query results to sort by the `population` column in ascending order.
What if we want to go the other way?

Well, you can sort results in **descending** order by tacking on the
`DESC` operator at the end of your `ORDER BY` clause:

``` markdown
SELECT
  column1,
  column2
FROM 
  someTable
ORDER BY
  column1 DESC;
```

You can also explicitly tell the query to sort results in ascending
order by using `ASC`. If you do not specify either `ASC` or `DESC`, the
database will sort in ascending order by default.

Starting from scratch, select all countries and arrange the results 
by population **in descending order**. Your results should only include the
columns `country`, `region`, and `population`.

```{sql db-exOrderBy2, exercise = TRUE, connection = "db_con", output.var="my_result", exercise.reveal_solution = FALSE}

```

```{sql db-exOrderBy2-solution}
SELECT 
  country, 
  region, 
  population
FROM 
  tblCountries
ORDER BY
  population DESC;
```

```{r db-exOrderBy2-check}
grade_this({
  .solution <- DBI::dbGetQuery(db_con, .solution_code)

  # # if (unique(.solution$year) == "2023") {
  # #   fail("Did you remember to select only data from the year 2023?")
  # # }

  pass_if_equal()
  fail()
})
  
```

### ORDER BY exercise 3

It's also possible to sort your data by multiple columns. To do so,
simply add more columns to the `ORDER BY` clause, separating them with
commas.

In this next exercise, select all countries and arrange the results 
(1) alphabetically by region **in ascending order** ⬇️, and 
(2) by population **in descending order** ⬆️.
Your result should only include the columns `country`, `region`, and
`population`.

```{sql db-exOrderBy3, exercise = TRUE, connection = "db_con", output.var="my_result", exercise.reveal_solution = FALSE}

```

```{sql db-exOrderBy3-solution}
SELECT 
  country, 
  region, 
  population
FROM 
  tblCountries
ORDER BY
  region ASC,
  population DESC;
```

```{r db-exOrderBy3-check}
grade_this({
  .solution <- DBI::dbGetQuery(db_con, .solution_code)

  # # if (unique(.solution$year) == "2023") {
  # #   fail("Did you remember to select only data from the year 2023?")
  # # }

  pass_if_equal()
  fail()
})
  
```

## SELECT DISTINCT

In data management terminology, the **grain** of a dataset (or
**granularity**) refers to the level of detail at which the data is
stored. For example, in the case of our population table, the data is
stored at a **grain** of "one row per country per year."

Suppose we want to return a list of the **distinct** countries in our
table. If we just write `SELECT country FROM tblPopulation`, this will
not produce the intended result because there are repeating values in
the `country` column.

Instead, we need to supplement our `SELECT` clause with the `DISTINCT`
operator:

``` markdown
SELECT DISTINCT
  column1
FROM 
  someTable;
```

### SELECT DISTINCT exercise 1

Retrieve a list of all distinct regions, and arrange them in ascending
alphabetical order.

```{sql db-exSelectDistinct1, exercise = TRUE, connection = "db_con", output.var="my_result", exercise.reveal_solution = FALSE}

```

```{sql db-exSelectDistinct1-solution}
SELECT DISTINCT
  region
FROM 
  tblCountries
ORDER BY
  region;
```

```{r db-exSelectDistinct1-check}
grade_this({
  .solution <- DBI::dbGetQuery(db_con, .solution_code)

  # # if (unique(.solution$year) == "2023") {
  # #   fail("Did you remember to select only data from the year 2023?")
  # # }

  pass_if_equal()
  fail()
})
  
```

### SELECT DISTINCT exercise 2

We can reference multiple columns inside a `SELECT DISTINCT` clause, but
it may not return the intended results, so be careful when doing this.

To illustrate what happens, write a query below that returns the
distinct values for both the `country` and `region` columns, and arrange
the results in ascending order for both columns:

```{sql db-exSelectDistinct2, exercise = TRUE, connection = "db_con", output.var="my_result", exercise.reveal_solution = FALSE}

```

```{sql db-exSelectDistinct2-solution}
SELECT DISTINCT
  country,
  region
FROM 
  tblCountries
ORDER BY
  country,
  region;
```

```{r db-exSelectDistinct2-check}
grade_this({
  .solution <- DBI::dbGetQuery(db_con, .solution_code)

  # # if (unique(.solution$year) == "2023") {
  # #   fail("Did you remember to select only data from the year 2023?")
  # # }

  pass_if_equal()
  fail()
})
  
```

### SELECT DISTINCT wrap-up

As you can see, using `SELECT DISTINCT` with multiple columns will
return every distinct combination of those columns. Make sure that's
what you intend!

## WHERE

It is almost always the case that we will need to remove data from our
query results through filtering. There are a few different ways to
filter data in SQL, but the most fundamental way is by employing the
`WHERE` clause to filter columns.

This clause gets added to your query after the `FROM` clause, and allows
you to use various operators for comparison and matching. For example,
returning all rows that equal a particular value is as easy as using an
equals sign (=).

``` markdown
SELECT
  column1,
  column2
FROM 
  someTable
WHERE
  column1 = 'someValue'
```

### WHERE exercise 1

Select all countries in the Middle East & North Africa region. Your result
should only include the columns `country`, `region`, and `population`.
Arrange the results in descending order by population.

```{sql db-exWhere1, exercise = TRUE, connection = "db_con", output.var="my_result", exercise.reveal_solution = FALSE}

```

```{sql db-exWhere1-solution}
SELECT 
  country, 
  region, 
  population
FROM 
  tblCountries
WHERE 
  region = 'Middle East & North Africa'
ORDER BY
  population DESC;
```

```{r db-exWhere1-check}
grade_this({
  .solution <- DBI::dbGetQuery(db_con, .solution_code)

  # # if (unique(.solution$year) == "2023") {
  # #   fail("Did you remember to select only data from the year 2023?")
  # # }

  pass_if_equal()
  fail()
})
  
```

### WHERE exercise 2

Of course, there are other operators we can use besides the equals sign
to run comparisons for filtering purposes. As you know from our
Practical SQL reading (pg 36), there are several more ways we can
compare and match data for the purpose of narrowing down our results.
Let's explore some of these in the next few exercises.

Retrieve a list of all distinct countries that are NOT Afghanistan,
arranged in ascending alphabetical order.

**NOTE: you can use either `<>` or `!=` to perform a not-equals
comparison.**

```{sql db-exWhere2, exercise = TRUE, connection = "db_con", output.var="my_result", exercise.reveal_solution = FALSE}

```

```{sql db-exWhere2-solution}
SELECT DISTINCT
  country
FROM 
  tblPopulation
WHERE 
  country <> 'Afghanistan'
ORDER BY
  country ASC
```

```{r db-exWhere2-check}
grade_this({
  .solution <- DBI::dbGetQuery(db_con, .solution_code)

  pass_if_equal()
  fail()
})
  
```

### WHERE exercise 3

When specifying multiple filter criteria in your `WHERE` clause, you
need to rely on the `AND` and `OR` operators to construct the logical
basis on which the database engine can determine which rows to keep or
remove.

With this mind, write a new query that retrieves a list of countries
that had a population greater than 150 million in the year 2000. Include
the columns `country`, `year`, and `population`.

NOTE: you do not need to use `DISTINCT` for this because your query
should return results at a grain of one row per country.

```{sql db-exWhere3, exercise = TRUE, connection = "db_con", output.var="my_result", exercise.reveal_solution = FALSE}

```

```{sql db-exWhere3-solution}
SELECT 
  country,
  year,
  population
FROM 
  tblPopulation
WHERE 
  population > 150000000
  AND year = 2000
```

```{r db-exWhere3-check}
grade_this({
  .solution <- DBI::dbGetQuery(db_con, .solution_code)
  
  pass_if_equal()
  fail()
})
  
```

### WHERE exercise 4

Let's take a look at the `OR` operator now, which allows us to return
rows that meet either Condition A or Condition B.

Using the `OR` operator inside your `WHERE` clause, retrieve a list of
countries with population figures from 2000 or 2001. Include the columns
`country`, `year`, and `population`.

```{sql db-exWhere4, exercise = TRUE, connection = "db_con", output.var="my_result", exercise.reveal_solution = FALSE}

```

```{sql db-exWhere4-solution}
SELECT 
  country,
  year,
  population
FROM 
  tblPopulation
WHERE 
  year = '2000' OR year = '2001'
```

```{r db-exWhere4-check}
grade_this({
  .solution <- DBI::dbGetQuery(db_con, .solution_code)
  
  pass_if_equal()
  fail()
})
  
```

### WHERE exercise 5

Using the `OR` operator can be tricky in any programming language, and
SQL is no exception. Thanks to the order of operations, combining `AND`
along with `OR` can create situations where you return more (or less)
data than you intend.

To mitigate this, we can place parentheses `(` and `)` around parts of
our query. This has the effect of telling the database engine to
evaluate the criteria contained within those parentheses separately.

For this next exercise, experiment with running the query with **and
without** the parentheses, and compare the results you get.

Write a query that returns the Brazilian population figures for 2014 and
2015. Include the columns `country`, `year`, and `population`.

```{sql db-exWhere5, exercise = TRUE, connection = "db_con", output.var="my_result", exercise.reveal_solution = FALSE}

```

```{sql db-exWhere5-solution}
SELECT 
  country,
  year,
  population
FROM 
  tblPopulation
WHERE 
  country = 'Brazil'
  AND (year = '2014' OR year = '2015')
```

```{r db-exWhere5-check}
grade_this({
  .solution <- DBI::dbGetQuery(db_con, .solution_code)
  
  pass_if_equal()
  fail()
})
  
```

### WHERE exercise 6

## SELECT \*

## Quiz
